A Generalized Sphere Theorem for Positively Curved Combinatorial 3-Manifolds

### 2013-06-14 paper/paper.tex
more writing, fixed some small problems with the graphs

\begin{abstract}
The generalized sphere theorem of Grove and Shiohama is an important result in the differential geometry of positively curved manifolds. It shows that if $M$ is a Riemannian manifold whose sectional curvature is everywhere greater than a positive constant and $M$ has diameter greater than half the maximum allowed by the Bonnet-Myers theorem, then $M$ must be homeomorphic to a sphere. In this paper, we present a novel discrete version of this theorem which applies to positively curved {\em combinatorial} 3-manifolds. That is, those with at most five tetrahedra incident at each edge, or equivalently, those with an angle deficit at each edge in the standard piecewise-linear metric. Such spaces have been studied previously and found to satisfy a beautiful discrete version of the Bonnet-Myers theorem. Here we prove the corresponding {\em discrete} generalized sphere theorem using a complete census of the positively curved 3-manifolds completed by Lutz and Sullivan.
\end{abstract}

@@

Just as for edges in an edge-path, the length of each hop and jump will be its length as a PL-path in the standard PL metric. Some Euclidean geometry tells us these lengths are \begin{equation} \label{eqn:hop_length} H = \frac{2}{3}\sqrt{2} \end{equation} and \begin{equation} \label{eqn:jump_length} J = \frac{1}{2}\sqrt{2} + \sqrt{3} \end{equation} respectively. We let $d(v,w)$ denote the \textbf{distance between vertices $v$ and $w$} obtained by minimizing over all paths containing edges as well as hops and jumps. Similarly, we let $diam(M)$ denote the \textbf{diameter of $M$} defined in terms of the distance function $d$. Let $\mathcal{P}$ denote the set of all the paths containing edges, hops or jumps.

Why are these paths relevant to this paper? In the classical setting, the generalized sphere theorem requires a manifold to have diameter more than half the maximum allowed by the Bonnet-Myers theorem, and this diameter bound cannot be improved. If we naively imitate the classical results in the discrete setting, from Theorem \ref{thm:discrete_BM} we would want to assume the bound $diam_1(M)>\frac{5}{2}$. Unfortunately, there are positively curved combinatorial 3-manifolds with edge-diameter three that are not homeomorphic to the 3-sphere. In particular, as we will see, $\mathbb{RP}^3$ has positively curved triangulations with edge-diameter three. However, if we instead use the finer-grained measure of diameter, $diam(M)$, the problem disappears and the discrete results mirror the classical ones exactly. That is, we can prove a discrete version of the generalized sphere theorem with a diameter bound exactly half the maximum allowed by the corresponding discrete Bonnet-Myers theorem.

@@

\begin{thm} Any connected positively curved combinatorial 3-manifold $M$ with $diam(M)>J$ is homeomorphic to the 3-sphere. This bound is sharp, and equal to half the maximum diameter which occurs for any such $M$.
\label{thm:discrete_GS}
\end{thm}

\noindent The proof of this theorem relies on the remarkable work of Lutz, Sullivan and Sulanke in \cite{Lutz07, LutzSul_unpub, sulanke2009isomorphism} who created a complete census of positively curved combinatorial 3-manifolds along with a list of their topological types, see \cite{Lutz_online_manifolds} and \cite{Lutz_online_topological_types}. This census contains 4787 manifolds, with the following topologies represented: the 3-sphere $S^3$, the real projective space $\mathbb{RP}^3$, the lens spaces $L(3,1)$ and $L(4,1)$, and the cube space $S^3/Q$. Note that all of these manifolds are homeomorphic to either a 3-sphere or to the quotient space of a 3-sphere under the action of a finite group.

A small Python program (available on request) was created to examine each manifold in the census and compute its diameter. See Figure \ref{fig:type_statistics} for a plot of the number of manifolds by topological type and diameter. As claimed, every manifold with diameter greater than $J$ is a 3-sphere. The corresponding data using the coarser-grained edge-diameter can be found in Figure \ref{fig:type_statistics_edges}. Note the existence of $3$-manifolds with edge-diameter greater than half the maximum allowed edge-diameter and which are {\em not} homeomorphic to the 3-sphere. In particular, there are positively curved combinatorial 3-manifolds homeomorphic to $\mathbb{RP}^3$, $L(3,1)$ and $L(4,1)$ which have edge-diameter three.

@@

\begin{figure}
    \begin{center}
    {
	    \setlength\fboxsep{15pt}
		\setlength\fboxrule{0pt}
		\fbox{\includegraphics[width=0.9\linewidth]{figures/split_statistics.png}}
    }
    \renewcommand{\arraystretch}{1.2}
	\begin{tabular} {| r | r | r | r | r | r | r |}	
		\hline
		Edge-Diameter & $S^{3}/Q$ & $L(4,1)$ & $L(3,1)$ & $RP^{3}$ & $S^{3}$ & Total \\
		\hline
		\hline
		$1$ &  &  &  &   &3    & 3    \\
		$2$ & 1&  &1 &17 &574  & 593  \\
		$3$ &  &1 &1 &5  &3498 & 3505 \\
		$4$ &  &  &  &   &675  & 675   \\
		$5$ &  &  &  &   &11   & 11   \\
		\hline
		Total&1 &1 &2 &22   &4761   &4787  \\
		\hline
	\end{tabular}
    \end{center}
    \caption{Number of connected positively curved combinatorial 3-manifolds by topology and edge-diameter, displayed graphically (above) and in table form (below). Note that when diameters are measured in edges, there are manifolds with diameter greater than half the maximum which are {\em not} 3-spheres. Results were computed using a 3-manifold census created by Lutz and Sullivan \cite{LutzSul_unpub}.}
	\label{fig:type_statistics_edges}
\end{figure}

\section{Discussion}
Why does adding additional paths above and beyond edge-paths restore the classical structure to our proof?

We should mention another notable advantage of expanding the space of paths to include hops and edges. In classical differential geometry, the starting point, the initial direction, and length completely determine a geodesic. There is no branching of geodesics. This property of fundamental importance to many proofs in Riemannian geometry. If we use only edge-paths, then in a positi`vely curved combinatorial 3-manifold it is possible to have two minimal paths with the same length and which begin by traversing the same edge, yet reach distinct destination vertices. That is, using only edge paths, minimal geodesics can ``branch''.

>>>>

\begin{abstract}
The generalized sphere theorem of Grove and Shiohama is an important result in the differential geometry of positively curved manifolds. It shows that if $M$ is a Riemannian manifold whose sectional curvature is everywhere greater than a positive constant and $M$ has diameter greater than half the maximum allowed by the Bonnet-Myers theorem, then $M$ must be homeomorphic to a sphere. In this paper, we present a novel discrete version of this theorem which applies to positively curved {\em combinatorial} 3-manifolds. That is, those with at most five tetrahedra incident at each edge, or equivalently, those with an angle deficit at each edge in the standard piecewise-linear metric. Such spaces have been studied previously and found to satisfy a beautiful discrete version of the Bonnet-Myers theorem. Here, we use a complete census of the positively curved 3-manifolds completed by Lutz and Sullivan to prove a corresponding {\em discrete} generalized sphere theorem.
\end{abstract}

@@

Just as for edges in an edge-path, the length of each hop and jump will be its length as a PL-path in the standard PL metric. Some Euclidean geometry tells us these lengths are \begin{equation} \label{eqn:hop_length} H = \frac{2}{3}\sqrt{2} \end{equation} and \begin{equation} \label{eqn:jump_length} J = \frac{1}{2}\sqrt{2} + \sqrt{3} \end{equation} respectively. We let $d(v,w)$ denote the \textbf{distance between vertices $v$ and $w$} obtained by minimizing over all paths containing edges as well as hops and jumps. Similarly, we let $diam(M)$ denote the \textbf{diameter of $M$} defined in terms of the distance function $d$. Finally, let $\mathcal{P}$ denote the set of all the paths containing edges, hops or jumps.

Why are these paths relevant to this paper? In the classical setting, the generalized sphere theorem requires a manifold to have diameter more than half the maximum allowed by the Bonnet-Myers theorem, and this diameter bound cannot be improved. If we naively imitate the classical results in the discrete setting, from Theorem \ref{thm:discrete_BM} we would want to assume the bound $diam_1(M)>\frac{5}{2}$. Right away we see a discrepancy between the discrete and classical results since this bound cannot possibly be sharp. Worse, the smallest weakening that could be sharp is $diam_1(M) \geq 3$ and this bound does not work. As we will see, there are positively curved combinatorial 3-manifolds $M$ with $diam_1(M)=3$ but which are {\em not} homeomorphic to the 3-sphere. However, if we instead use the finer-grained measure of diameter, $diam(M)$, the problem disappears and the discrete results mirror the classical smooth ones exactly. That is, we can prove a discrete version of the generalized sphere theorem with a diameter bound exactly half the maximum allowed by the corresponding discrete Bonnet-Myers theorem.

@@

\begin{thm} Let $M$ be a connected, positively curved, combinatorial 3-manifold. If $M$ has $diam(M)>J$ then it is homeomorphic to a 3-sphere. Moreover, this bound is sharp and equal to half the maximum diameter which occurs for any such $M$.
\label{thm:discrete_GS}
\end{thm}

\noindent The proof of this theorem relies on the remarkable work of Lutz, Sullivan and Sulanke in \cite{Lutz07, LutzSul_unpub, sulanke2009isomorphism} who created a complete census of positively curved combinatorial 3-manifolds along with a list of their topological types. See \cite{Lutz_online_manifolds} and \cite{Lutz_online_topological_types}. This census contains 4787 manifolds, with the following topologies represented: the 3-sphere $S^3$, the real projective space $\mathbb{RP}^3$, the lens spaces $L(3,1)$ and $L(4,1)$, and the cube space $S^3/Q$. Note that all of these manifolds are homeomorphic to either a 3-sphere or to the quotient space of a 3-sphere under the action of a finite group.

A small Python program (available on request) was created to examine each manifold in the census and compute its diameter and edge-diameter. See Figure \ref{fig:type_statistics} for a plot of the number of manifolds by topological type and diameter. As claimed, every manifold with diameter greater than $J$ is a 3-sphere. The corresponding data using the coarser-grained edge-diameter can be found in Figure \ref{fig:type_statistics_edges}. Note the existence of $3$-manifolds with edge-diameter greater than half the maximum allowed edge-diameter and which are {\em not} homeomorphic to the 3-sphere.

If one absolutely insists on results stated in terms of edge-diameter then the census does give us the following result:

\begin{thm} If $M$ is a connected, positively curved, combinatorial 3-manifold satisfying the bound $diam_1(M) > 3$ then $M$ is homeomorphic to the 3-sphere. Moreover, this bound is sharp.
\end{thm}

@@

\begin{figure}
    \begin{center}
    {
	    \setlength\fboxsep{15pt}
		\setlength\fboxrule{0pt}
		\fbox{\includegraphics[width=0.9\linewidth]{figures/split_edge_diameters.png}}
    }
    \renewcommand{\arraystretch}{1.2}
	\begin{tabular} {| r | r | r | r | r | r | r |}	
		\hline
		Edge-Diameter & $S^{3}/Q$ & $L(4,1)$ & $L(3,1)$ & $RP^{3}$ & $S^{3}$ & Total \\
		\hline
		\hline
		$1$ &  &  &  &   &3    & 3    \\
		$2$ & 1&  &1 &17 &574  & 593  \\
		$3$ &  &1 &1 &5  &3498 & 3505 \\
		$4$ &  &  &  &   &675  & 675   \\
		$5$ &  &  &  &   &11   & 11   \\
		\hline
		Total&1 &1 &2 &22   &4761   &4787  \\
		\hline
	\end{tabular}
    \end{center}
    \caption{Number of connected positively curved combinatorial 3-manifolds by topology and edge-diameter, displayed graphically (above) and in table form (below). Note that when diameters are measured in edges, there are manifolds with diameter greater than half the maximum which are {\em not} 3-spheres. Results were computed using a 3-manifold census created by Lutz and Sullivan \cite{LutzSul_unpub}.}
	\label{fig:type_statistics_edges}
\end{figure}

\section{Discussion}
We should mention another notable advantage of expanding the space of paths to include hops and edges. In classical differential geometry, the starting point, the initial direction, and length completely determine a geodesic. There is no branching of geodesics. This property of fundamental importance to many proofs in Riemannian geometry. If we use only edge-paths, then in a positi`vely curved combinatorial 3-manifold it is possible to have two minimal paths with the same length and which begin by traversing the same edge, yet reach distinct destination vertices. That is, using only edge paths, minimal geodesics can ``branch''.
